\documentclass[a4paper, 11pt,twoside,openright]{book}

\input{structure}



\title{Momentum space correlations in the depletion of weakly interacting lattice Bose gases}

\author{Antoine Ténart}

\institute{Institut d'Optique Graduate School}
\doctoralschool{Ondes et Matière}{572}{EDOM}
\specialty{Physique}
\date{TBD}
\NNT{TBD}

%\jurymember{1}{Prénom NOM}{Titre, \'etablissement}{Président}
%% Faut-il mettre le grade des personnes ?)
% \jurymember{4}{Jean-François CLEMENT}{Maître de conférences au Laboratoire de Physique des Lasers, Atomes et Molécules - Université de Lille}{Examinateur}
% \jurymember{3}{Robin KAISER}{Directeur de recherche à l'Institut de Physique de Nice - Université Côte d'Azur}{Rapporteur}
% \jurymember{2}{Dominique DELANDE}{Directeur de recherche au Laboratoire Kastler Brossel - Sorbonne Université}{Rapporteur}
% \jurymember{1}{Laurence PRUVOST}{Directrice de recherche au Laboratoire Aimé Cotton - Université Paris Saclay}{Présidente}
% \jurymember{6}{Alain ASPECT}{Directeur de recherche émérite au Laboratoire Charles Fabry - Institut d'Optique Graduate School - Université Paris Saclay}{Membre invité}
% \jurymember{5}{Vincent JOSSE}{Maître de conférences au Laboratoire Charles Fabry - Institut d'Optique Graduate School - Université Paris Saclay}{Directeur de thèse}

\frtitle{Corrélations en impulsion dans la déplétion de gaz de Bose faiblement interagissant}

\entitle{Momentum space correlations in the depletion of weakly interacting lattice Bose gases}

\frabstract{Ce travail de thèse est centré sur l'étude d'un exemple emblématique de système quantique en interaction, le condensat de Bose-Einstein faiblement interagissant. A température nulle, la théorie de Bogoliubov prévoit que l'effet conjoint des interactions et des fluctuations quantiques retire une fraction des atomes du condensat. Cette fraction est nommée déplétion quantique. De manière intéressante, la déplétion quantique est exclusivement constituée de paires d'atomes de vitesses opposées et donc corrélés en impulsion. Bien que cette prédiction date de 60 ans, ce signal de corrélation n'avait jamais été observé auparavant. Ce manuscrit rapporte la première observation de corrélations entre impulsions opposées dans la déplétion d'un condensat de Bose-Einstein faiblement interagissant. Pour ce faire, nous produisons des condensats de Bose-Einstein d'Hélium-4 métastable chargés dans des réseaux optiques. L'utilisation d'Hélium métastable rend possible la détection d'atomes individuels en 3D après un long temps de vol, aspect essentiel à la mesure d'un signal de corrélation entre particules individuelles. Après avoir exposé plusieurs résultats visant à démontrer que notre technique de détection mesure fidèlement l'impulsion d'atomes individuels, nous présentons nos signaux de corrélations dont nous étudions les principales caractéristiques (amplitude, largeur, évolution avec la température) pour mettre en valeur leur nature quantique. Au delà de confirmer une prédiction théorique, ces résultats constituent un premier pas dans l'étude des systèmes quantiques à N corps interagissant à l'équilibre au travers des corrélations en impulsion et ouvrent la voie à des études de corrélations plus complexes. Pour finir, nous présentons des résultats préliminaires sur un sujet différent, la mesure du contact de Tan via la densité en impulsion dans des gaz de Bose unidimensionnels. 



}

\enabstract{This thesis is focused on the study of the emblematic example of an interacting quantum system, the weakly-interacting Bose-Einstein condensate. At zero temperature, a fraction of the atoms is removed from the condensate through the interplay between the interactions and the quantum fluctuations. This fraction is called the quantum depletion. Interestingly, the quantum depletion is entirely made of pairs of atoms with opposite speeds, put in another way momentum correlated pairs. While this prediction is 60 years old, this correlation signal had never been observed before. This manuscript reports the first observation of opposite momentum correlations in the depletion of a weakly-interacting Bose-Einstein condensate. To do so, we produce metastable Helium-4 condensates loaded in optical lattices. Using metastable Helium makes possible the detection of individual atoms in 3D after a long time-of-flight, an essential aspect to measure a correlation signal between individual particles. After exposing several results proving that our detection technique faithfully measures the the momentum of individual atoms, we present our correlation signals of which we study the main characteristics (amplitude, width, evolution with temperature) to highlight their quantum nature. More than confirming a theoretical prediction, these results constitute a first step in the study of interacting many-body quantum systems at equilibrium and open the way to the study of more complex correlation patterns. In addition, we also present some preliminary results on a different project, namely the measure of Tan's contact using the momentum density in 1D Bose gases.


}

\frkeywords{Détection d'atomes uniques, Réseau optique, Corrélation en vitesse, Gaz quantique, Problème à N corps, Déplétion quantique}
\enkeywords{Single atom detection, Optical lattice, Momentum correlation, Quantum gas, Many-body problem, Quantum depletion}



\begin{document}
\maketitle{}

\usechapterimagetrue
\setlength{\parskip}{\parskipnew}

{\hypersetup{linkcolor=black}
\tableofcontents}
\setlength{\parskip}{\parskipnew}





\input{Chapters/Intro}

\input{Chapters/Chapter1_Quantum_Correlations}

\input{Chapters/Chapter2_BH_Model}

\input{Chapters/Chapter3_MomentumDistritbution_lattice_gas}

\input{Chapters/Chapter4_kmk}

\input{Chapters/Chapter5_1D_gas}

\input{Chapters/Conclusion}


\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{\listfigurename}


\fancyhead[LO]{\sffamily\bfseries List of Figures} % Print the nearest section name on the left side of odd pages
\fancyhead[RE]{\sffamily\bfseries List of Figures} % Print the current chapter name on the right side of even pages
\listoffigures

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{\listtablename}
\listoftables

\cleardoublepage
\phantomsection
\chapter*{Publications}
\label{chap:publications}
\addcontentsline{toc}{chapter}{\nameref{chap:publications}}

\begin{itemize}
    \item C. Carcy, H. Cayla, A. Tenart, A. Aspect, M. Mancini, and D. Clément. \textbf{Momentum-space atom correlations in a mott insulator}. Physical Review X, 9(4):041028, 2019. Used as reference \cite{carcy2019momentum} in this manuscript.
    \vspace{0.5cm}
    \item A. Tenart, C. Carcy, H. Cayla, T. Bourdel, M. Mancini, and D. Clément. \textbf{Two-body collisions in the time-of-flight dynamics of lattice bose superfluids}. Physical Review Research, 2(1):013017, 2020. Used as reference \cite{tenart2020two} in this manuscript.
    \vspace{0.5cm}
    \item H. Cayla, S. Butera, C. Carcy, A. Tenart, G. Hercé, M. Mancini, A. Aspect, I. Carusotto, and D. Clément. \textbf{ Hanbury-brown and twiss bunching of phonons and of the quantum depletion in a strongly-interacting bose gas}. Physical Review Letters, 125:165301, 2020. Used as reference \cite{cayla2020} in this manuscript.
    \vspace{0.5cm}
    \item C. Carcy, G. Hercé, A. Tenart, T. Roscilde, and D. Clément. \textbf{Certifying the adiabatic preparation of ultracold lattice bosons in the vicinity of the mott transition}. Physical Review Letters, 126(4):045301, 2021. Used as reference \cite{carcy2021} in this manuscript.
    \vspace{0.5cm}
    \item G. Hercé, C. Carcy, A. Tenart, J.-P. Bureik, A. Dareau, D. Clément, and T. Roscilde. \textbf{Studying the low-entropy mott transition of bosons in a three-dimensional optical lattice by measuring the full momentum-space density}. Physical Review A, 104, 2021. Used as reference \cite{herce2021studying} in this manuscript.
    \vspace{0.5cm}
    \item A. Tenart, G. Hercé, J.-P. Bureik, A. Dareau, and D. Clément. \textbf{Observation of pairs of atoms at opposite momenta in an equilibrium interacting bose gas}. arXiv preprint arXiv:2105.05664, 2021. Accepted for publication in Nature Physics. Used as reference \cite{tenart2021observation} in this manuscript. 
\end{itemize}




\setlength{\parskip}{\parskipinitial}
%\usechapterimagefalse



\addcontentsline{toc}{chapter}{Bibliography}
\fancyhead[LO,RE]{\sffamily\normalsize\bfseries Bibliography}

\bibliography{sample}



\end{document}
